---
title: "Decision Making"
subtitle: ""
author: "Tom Bruning"
date: "`r Sys.Date()`"
output:
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex

---

```{r setup, include=FALSE}
library(tufte)
library(knitr)
library(tibble)
library(tidyr)
library(dplyr)

# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

# Steps in Decision Making  

* List Alternative Courses of Action   
    * Choices or actions  
* List Uncertain Events  
    * Possible events or outcomes  
* Determine ‘Payoffs’  
    * Associate a Payoff with Each Choice/Event combination  
* Adopt Decision Criteria  
    * Evaluate Criteria for Selecting the Best Course of Action  
    
## A Payoff Table  
A payoff table shows alternatives, states of nature, and payoffs

```{r, echo=FALSE}

a <- tibble(c("Strong Economy", "Stable Economy", "Weak Economy"))
b <- tibble(c(200, 50, -120))
c <- tibble(c(90, 120, -30))
d <- tibble(c(40, 30, 20))
e <- bind_cols(a,b,c,d)
kable(e, col.names = c("States of Nature", "Large Factory", "Average Factory", "Small Factory"), caption = "Payoff Table \nProfit in $1000s")
```

## Opportunity Loss  
Opportunity loss is the difference between an actual payoff for an action and the highest possible payoff, given a particular event

The action “Average factory” has payoff 90 for “Strong Economy”.  Given “Strong Economy”, the choice of “Large factory” would have given a payoff of 200, or 110 higher.  *Opportunity loss = 110 for this cell.*

```{r, echo=FALSE}

a <- tibble(c("Strong Economy", "Stable Economy", "Weak Economy"))
b <- tibble(c(0, 70, 140))
c <- tibble(c(110, 0, 50))
d <- tibble(c(160, 90, 0))
e <- bind_cols(a,b,c,d)
kable(e, col.names = c("States of Nature", "Large Factory", "Average Factory", "Small Factory"), caption = "Opportunity Loss Table Opportunity Loss in $1000s")
```

## Decision Criteria  

* Maximax  
    * An optimistic decision criteria  
* Maximin  
    * A pessimistic decision criteria  
* Expected Monetary Value (EMV(j))   
    * The expected profit for taking action j  
* Expected Opportunity Loss (EOL(j))  
    * The expected opportunity loss for taking action j  
* Expected Value of Perfect Information (EVPI)  
    * The expected opportunity loss from the best decision  
* Return to Bayes

## Maximax Solution

* Maximum payoff for **Large Factory** is 200  
* Maximum payoff for **Average Factory** is 120  
* Maximum payoff for **Small Factory** is 40  

Maximax Decision:  Build the Large Factory because 200 is the maximum  
## Maximin Solution  

* Minimum payoff for **Large Factory** is -120  
* Minimum payoff for **Average Factory** is -30  
* Minimum payoff for **Small Factory* is 20  

Maximin Decision:  Build the Small Factory because 20 is the maximum

## Expected Monetary Value Solution 

Goal:  Maximize expected value  
The expected monetary value is the weighted average payoff, given specified probabilities for each event.  
\begin{equation}
  EMV_j = \sum_{i=1}^N x_{ij} P(X=x_i)
\end{equation}
Where:  
  $EMV_j$ = expected monetary value of action j  
	$x_{ij}$ = payoff for action j when event i occurs  
  $P_i$ = probability of event i    

```{r, echo=FALSE}

a <- tibble(c("Strong Economy(.3)", "Stable Economy(.5)", "Weak Economy(.2)", "EMV"))
b <- tibble(c(200, 500, -120, 61))
c <- tibble(c(90, 120, -30, 81))
d <- tibble(c(40, 30, 20, 31))
e <- bind_cols(a,b,c,d)
kable(e, col.names = c("States of Nature", "Large Factory", "Average Factory", "Small Factory"), caption = "Payoff Table \nProfit in $1000s")
```

Example:  
EMV (Average factory) =   (90)(.3) + (120)(.5) + (-30)(.2) =  81 

Maximize expected value by choosing **Average factory**  

## Expected Opportunity Loss Solution  

Goal:  Minimize expected opportunity loss
The expected opportunity loss is the weighted average loss, given specified probabilities for each event
\begin{equation}
  EOL_j = \sum_{i=1}^N Lx_{ij} P(X=x_i)
\end{equation}
Where:       
    $EOL_j$ = expected opportunity loss of action j  
    $L_{ij}$ = opportunity loss for action j when event i occurs  
    $P_i$ = probability of event i  

## Expected Opportunity Loss Solution
Goal:  Minimize expected opportunity loss

```{r, echo=FALSE}

a <- tibble(c("Strong Economy(.3)", "Stable Economy(.5)", "Weak Economy(.2", "EOL"))
b <- tibble(c(0, 70, 140, 63))
c <- tibble(c(110, 0, 50,43))
d <- tibble(c(160, 90, 0, 93))
e <- bind_cols(a,b,c,d)
kable(e, col.names = c("States of Nature", "Large Factory", "Average Factory", "Small Factory"), caption = "Opportunity Loss Table Opportunity Loss in $1000s")
```

Example:
EOL (Large factory) = 0(.3) + 70(.5) + (140)(.2) = 63
Minimize expected op. loss by choosing **Average factory** 

## Expected Opportunity Loss vs. Expected Monetary Value  

* The Expected Monetary Value (EMV) and the Expected Opportunity Loss (EOL) criteria are equivalent.  

* Note that in this example the expected monetary value solution and the expected opportunity loss solution both led to the choice of the average size factory.  

## Value of Information 

Expected Value of Perfect Information, EVPI  


**EVPI** = Expected profit under certainty	– expected monetary value of the best alternative  

(EVPI is equal to the expected opportunity loss from the best decision)


## Expected Profit Under Certainty
Expected profit under certainty = expected value of the best decision, *given perfect information*

Examples:   

* Best decision given “Strong Economy” is “Large factory” (200)
* Best decision given "Stable Economy"" is "Average factory" (120)
* Best decision given "Weak Economy" is "Small factory" (20)  

Now weight these outcomes with their probabilities to find the expected value.  
200(.3)+120(.5)+20(.2) = 124, which is the expected profit under certainty

## Value of Information Solution (EVPI) 

EVPI = Expected profit under certainty – Expected monetary value of the best decision  
Recall:      

* Expected profit under certainty = 124  
* EMV is maximized by choosing “Average factory”, where EMV = 81  
	  
so: 
* EVPI 	=  124 – 81 	=  43  
(EVPI is the maximum you would be willing to spend to obtain perfect information)

## Forecasting (accounting for randomness)

Suppose you are a district manager for a retail chain.  You have three stores in your district that have averaged the same store sales over the past three years of $10,000,000.  You know (from taking this class) that there is are random factors that are reflected in these averages. 
The store's past year sales were \$10,500,000, \$9,500,000, and \$10,000,000.
Your regional manager demands a 10% increase in sales from your district for the following year.  How best to allocate that $3,000,000 increase?

Since tha stores have the same average sales (for the past 3 years) you could increase their quota by $1,000,000. You could increase each store sales by 10%, or you could take randomness into account and factor that randomness into the new quotas.  
The following table summarizes these options.

```{r, echo=FALSE}

a <- tibble(c("3 yr. Average Sales", "Last Year Sales", "10% increase", "10% increase w/randomness"))
b <- tibble(c("$30,000,000","$30,000,000","$30,000,000","$30,000,000"))
c <- tibble(c("$10,000,000","$10,500,000","$11,550,000","$11,275,000"))
d <- tibble(c("$10,000,000","$9,500,000","$10,450,000","$10,725,000"))
e <- tibble(c("$10,000,000","$10,000,000","$11,000,000","$11,000,000"))
f <- bind_cols(a,b,c,d,e)
kable(f, col.names = c(" ", "District", "Store 1", "Store 2", "Store 3"), caption = "Summary")
```
  

