---
title: "Stat Meth Business  (33:136:385)"
author: "Tom Bruning"
date: "Sept 4, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructor

Thomas Bruning   
email: tom.bruning@gmail.com   
cell: 201.204.0585

**Office Hours:** by appointment

# Resources  

## Textbook:  
**Title:** *Basic Business Statistics: Concepts and Applications*  
**Authors:** Berenson, Levine,Krebbel  
**Publisher:** Pearson  
**ISBN-10 0132168383**  

## Computer Resources    
**Title:** Microsoft Excel, or equivalent  
**Additional Resources:** Data Analysis plugin  

## Prerequisites  
The following prerequisites are **essential** for successful learning and understanding of this course:    
* An introductory course in probability and statistics such as 960:285     
* Basic college algebra   
* Working knowledge of Excel spreadsheets

## Course Overview   
This course will review the basic statistical methods essential for doing business analysis, in particular when outcomes are uncertain and future events are subject to chance variation.  We will study basics of data analysis using computer software.

## Course Web site information and discussion   
### Blackboard
All class discussions, assignments, and materials will be posted to Blackboard   
Please review this site 2-3 times per week for further information.

# Homeworks, Exams, and Grading Policy   


## Homework (10% of final grade)  

I will post approximately 10 homework assignments during the semester.  You are responsible for completing the work, on time.  Statistics is not a *spectator sport* , it is a *participatory sport*.  I have it on good authority, (10+ years of teaching statistics experience) if you do not do the work you **will** not understand and succeed in this class.  If you miss two (2) homeworks you will receive an **F** for this class.  This is true regardless of your grades on the other homework assignment and your grades on the midterm and final.

## Midterm (40% of final grade)

The midterm is a full period test that is scheduled during the 7th week of class.

## Final (50% of final grade)

Schedule TBD   


## Topic Schedule
```{r grade1, echo=FALSE}
syllabus <- read.csv("./385_Syllabus.csv")
knitr::kable(syllabus, caption = "Tentative Topics and Dates")  
```

